Double Robust, Flexible Adjustment Methods for Causal Inference: An Overview and an Evaluation

### 2025-05-29 draft/ms-adj-draft.Rmd
revision

abstract: "Double robust methods for flexible covariate adjustment in causal inference have proliferated in recent years. Despite their apparent advantages, these methods are rarely used by social scientists. It is also unclear whether these methods actually outperform more traditional methods in finite samples. This paper has two aims: It is a guide to some of the latest methods in double robust, flexible covariate adjustment using machine learning, and it compares these methods to more traditional statistical methods and flexible \"single robust\" methods. It does this by using both simulated data where the treatment effect estimate is known, and then by replicating some prominent articles in sociology that use simpler methods. Double robust methods covered include Augmented Inverse Probability Weighting (AIPW), Targeted Maximum Likelihood Estimation (TMLE), and Double/Debiased Machine Learning (DML). Results suggest that some of these methods do outperform traditional methods in a wide range of simulations, but only slightly. In particular, the top performers are TMLE and AIPW in conjunction with flexible machine learning estimators. But G-computation with the same flexible estimators obtains almost identical results, and simple regression methods have only slightly higher bias and are much more computationally efficient. In the article replications, the application of double robust methods substantively changes some, but not all, of the results, highlighting the importance of comparing performance from multiple estimators."

@@

library(SuperLearner)
library(broom)
library(knitr)
library(kableExtra)
library(here)
library(patchwork)
library(haven)
library(tidyverse)

@@

@@

In causal inference, functional form misspecification of underlying models can bias estimates of treatment effects [@hernan_2020_causal; @morgan_2015_counterfactuals]. There have been two important developments that attempt to overcome this. First, methodologists have developed machine learning methods that allow greater flexibility in estimation, adjusting for covariates in data-driven, complex ways [@balzer_2021_invited; @brand_2023_recent]. The second development is double robust methods [@bang_2005_doubly; @kang_2007_demystifying], which estimate two models: one for treatment exposure and another for the outcome. These models are robust to misspecification of either one of these "nuisance" models.

@@

Second, this paper evaluates these methods. They are tested on simulations from @dorie_2019_automated, which cover a range of data-generating processes where ignorability holds -- i.e., there are no unmeasured confounders. In the simulations, double robust methods are compared to "single robust" methods (i.e., ones with one nuisance model). These include traditional or simpler statistical methods commonly used by social scientists: ordinary least squares (OLS) regression, matching on propensity scores estimated from logistic regression (PSM), and inverse probability weighting (IPW). They are also compared to two more flexible methods that may overcome the misspecification issues that double robust methods aim to overcome: G-computation and the Lin estimator [@lin_2013_agnostic]. Lastly, I use two of the double robust methods to replicate analyses from three recent *American Sociological Review* (*ASR*) papers that use OLS or logistic regression to estimate causal effects.

Results from simulations where ignorability holds show that some double robust methods outperform traditional statistical methods, but only slightly. AIPW and TMLE perform the best, at least when used in conjunction with flexible machine learning algorithms, while DML (in its partially linear form) does slightly worse than traditional methods. G-computation with flexible machine learning performs as well as the lowest-error double robust methods. Despite their lower average error, the computation times of these flexible methods are high. With its still relatively low error and much faster computation time, OLS remains a sensible choice as an estimator, and the Lin estimator -- which is as quick as standard OLS regression -- performs only slightly worse than the best double robust methods. In the *ASR* replications, double robust methods provide similar coefficients for most, but not all, of the analyses. In some cases, statistical significance is lost, and in a minority of cases the coefficient sign changes. These replications highlight the importance of testing the robustness of quantitative results by comparing multiple methods and estimators.

The next section reviews literature on and motivation for double robust methods. This is followed by an introduction of notation and assumptions, while the following section gives an overview of the double robust methods covered in this paper. I next present results of the simulations and replications, then conclude. In the Appendix, I provide simple R code to implement the three double robust methods.

@@

# Motivation

@@

Although previous work has compared different methods for covariate adjustment, there has not yet been a thorough comparison and evaluation of the recent and popular double robust methods of AIPW, TMLE, and DML for a social scientist audience.

@@

# Overview of Techniques  
Each of the methods reviewed in this paper can be used with a variety of estimation techniques, including both traditional statistical methods and flexible machine learning methods. Each involves a model for the outcome and another for the treatment exposure, but choice of estimator for these two models is left to the discretion of the user. Double robust methods are distinct in the ways these estimated models relate and are combined into a final estimate of the desired estimand.

This section provides some intuition for the mathematical theory behind each method. In the Appendix, I provide R code to simply implement these methods. This is the code used to evaluate these methods later in this paper, and it represents basic implementations of these methods. However, for researchers wishing to put these methods into practice, using a dedicated R packages for each method is probably a better idea. These packages have many flexible options, such as accounting for complex survey design, targeting estimands besides the ATE, and integrating with a variety of estimation techniques.

@@

Extending and improving previous double robust methods, @vanderlaan_2006_targeted first proposed TMLE using a parametric framework and the efficient influence curve [@hines_2022_demystifying] to obtain estimates and standard errors. Mark van der Laan has gone on to collaborate on both a gentle introduction [@gruber_2009_targeted], two textbooks [@vanderlaan_2011_targeted; @vanderlaan_2018_targeted], and an R package [@gruber_2012_tmle] for implementing the method. @schuler_2017_targeted and @luque-fernandez_2018_targeted provide introductions for epidemiologists.

@@

In the next step, we estimate fluctuation parameters $\epsilon = (\epsilon_0, \epsilon_1)$ through maximum likelihood of the following logistic regression with fixed intercept $\text{logit}(\mu_{di})$: 
<!-- $$E(Y=1 \mid D, \mathbf X) = \text{logit}(\mu_{di}) + \epsilon H_{di}$$ -->
$$\text{logit}[E(Y=1 \mid D, \mathbf X)] = \text{logit}(\hat \mu_{di}) + \epsilon_0 H_{0i} + \epsilon_1 H_{1i}$$

@@

For each individual in the sample, this estimator calculates two quantities:

- The treated potential outcome

@@

- The control potential outcome

\begin{equation}
\hat Y_{0i} = \frac{(1-D_i)(Y_i - \hat \mu_0 (X_i))}{1-\hat \pi(X_i)} + \hat \mu_0(X_i)
(\#eq:control-aipw)
\end{equation}

Let's focus on the treated model, Equation \@ref(eq:treated-aipw). First, assume that the outcome model $\mu_1(X_i)$ is *correctly* specified and the exposure model $\pi(X_i)$ is *incorrectly* specified. Let's also assume (for now) that we're dealing with a treated unit, i.e. $D_i = 1$. Then
$$E[\hat \mu_1 (X_i)] = E[Y_1 \mid X_i].$$
This means that
$$E[Y_i - \hat \mu_1(X_i)] = 0,$$
and hence
$$E[\hat Y_{1i}] = 0 + \hat \mu_1(X_i) = \hat \mu_1(X_i).$$
So the model relies *only* on the outcome model. The incorrectly specified exposure model completely disappears from the equation. If we're dealing with a control unit ($D_i=0$), we get the same result:
$$\hat Y_{1i} = \frac{0(Y_i - \hat \mu_1 (X_i))}{\hat \pi (X_i)} + \hat \mu_1(X_i) = \hat \mu_1(X_i).$$

Now, what if the *exposure* model $\pi(X_i)$ is correctly specified and the outcome model $\mu_1(X)$ is incorrect? First, we rewrite the estimator for the treated outcome:
\begin{align}
\hat Y_{1i}& = \frac{D_i(Y_i - \hat \mu_1 (X_i))}{\hat \pi (X_i)} + \hat \mu_1(X_i) \nonumber \\
&= \frac{D_iY_i}{\hat \pi (X_i)} - \frac{D_i\hat \mu_1 (X_i)}{\hat \pi (X_i)} + \frac{\hat \pi (X_i)\hat \mu_1(X_i)}{\hat \pi (X_i)} \nonumber \\
& = \frac{D_iY_i}{\hat \pi (X_i)} - \left( \frac{D_i - \hat \pi(X_i)}{\hat \pi (X_i)}\right) \hat \mu_1(X_i). 
(\#eq:exposure-correct)
\end{align}
Since the exposure model is correctly specified, we have $D_i = \hat \pi(X_i)$ on average, so
$$E[D_i - \hat \pi(X_i)] = 0.$$
This means that the second term in Equation \@ref(eq:exposure-correct) is 0, so
$$E[\hat Y_{1i}]= E \left [ \frac{D_iY_i}{\hat \pi (X_i)}\right].$$
This shows that when the exposure model is correct, then the estimator depends *only* on the exposure model. We can make similar arguments for the control model for $\hat Y_{0i}$ in Equation \@ref(eq:control-aipw).

This demonstration shows that this estimator achieves double robustness: The estimator is robust to misspecification of either the exposure or the outcome model (but not both). The other two double robust methods considered in this paper can be shown to have the same property, but proving so is more complicated.

@@

```{r dml-package, eval = F}
library(DoubleML)
library(mlr3)
library(mlr3pipelines)
library(mlr3learners)
library(mlr3extralearners)

@@

In 2016, the Atlantic Causal Inference Conference hosted a competition for causal inference methods that adjust on observables. @dorie_2019_automated published the results of this competition, along with the data used. Below, I test double robust methods on the 20 data sets used for the "do-it-yourself" part of the competition. The data represent a hypothetical twins study investigating the impact of birth weight on IQ. The data have 4,802 observations and 52 covariates. The authors of the study specify a different data generating process for the potential outcomes in each data set. In all cases, ignorability holds (all potential confounders are observed), but the authors vary the following:

- degree of nonlinearity
- percentage of treated
- overlap for the treatment group
- alignment (correspondence in variables used to generate the assignment mechanism and the response surface)
- treatment effect heterogeneity

@@

The second set of single robust methods are two-model methods, which estimate separate models for treated and untreated units. In theory these could solve the misspecification problem that double robust methods are meant to solve, but they could still suffer from regularization bias. G-computation [@robins_1986_new; @snowden_2011_implementation] uses some estimation technique to predict outcomes under treatment and control for each unit in the dataset. The ATE estimate is the difference in the average prediction under treatment and the average prediction under control. The second two-model method is the Lin estimator [@lin_2013_agnostic]. This method aims to solve issues with the bias induced by OLS regression in a randomization framework by interacting the treatment indicator with mean-centered covariates. @hazlett_2024_understanding show that this method is equivalent to estimating two separate OLS regression models for the treated and control units.

@@

The 20 data sets used here cover a range of these attributes; see the supplemental material from @dorie_2019_automated for details. I use 10 simulations of each data set, resulting in 200 data sets. I then calculate bias, percent bias (the estimator's bias as a percentage of its standard error), root mean squared error (rmse), and median absolute error (mae). I also present the number of datasets for which the method fails and the median computation time for each data set, in seconds.^[Simulations were run on a 2020 MacBook Pro laptop computer with a 2 GHz Quad-Core Intel Core i5 Processor and 16 GB of memory.] In the main text I present average bias and RMSE, while the Supplementary Material contains tables with full results.

@@

```{r dorie-results-bias,  fig.cap = 'Bias of Monte Carlo simulations using the first 20 datasets from Dorie et al. (2019), 10 replications each.'}
addline_format <- function(x,...){
    gsub('\\s','\n',x)
}

@@

```{r dorie-results-rmse, fig.cap = 'Root mean squared error and bias for Monte Carlo simulations using the first 20 datasets from Dorie et al. (2019), 10 replications each. Values greater in absolute value than 4 are plotted at 4 and labeled with their actual value.'}

@@

```{r dorie-linear-bias,  fig.cap = 'Bias of Monte Carlo simulations using the two datasets from Dorie et al. (2019), with linear data generating processes, 100 replications each ("linear").'}
sim_results %>% 
  filter(set == 'linear') %>%
  mutate(method_estimator = str_replace(method_estimator, 'superlearner', 'superl.')) %>%
  mutate(method_estimator = factor(method_estimator, levels = unique({.$method_estimator}))) %>%
  #filter(method %in% c('ols', 'psm', 'aipw')) %>%
  ggplot(aes(x = method_estimator, y = bias)) +
  geom_jitter(alpha = .2, width = .3, aes(color = method)) +
  geom_boxplot(alpha = .5, outlier.alpha = 0) +
  geom_hline(yintercept = 0) +
  #facet_wrap(~method, scales = 'free_x') +
  labs(x = '', title = 'Linear DGPs') +
  scale_y_continuous(trans=ggallin::pseudolog10_trans, breaks = c(-20, -10, -5, -1, 0, 1, 5)) +
  # scale_x_discrete(labels=function(x){sub(", ", "\n", x)}) +
  theme(axis.text.x=element_text(angle=90, hjust=0.95,vjust=0.2),
        legend.position = 'none')

@@

bind_rows(dgp_list) %>%
  select(-c(`Fourth-lowest`:`Sixth-lowest`)) %>%
  kableExtra::kable(booktabs = T, 
                    digits = 3,
                    linesep = '',
                    caption = 'Data generating process: Three lowest RMSE methods by DGP for Monte Carlo simulations using the first 20 datasets from Dorie et al. (2019), 10 replications each.')
```

```{r dgp-2}
bind_rows(dgp_list) %>%
  select(-c(`Lowest RMSE`:`Third-lowest`)) %>%
  kableExtra::kable(booktabs = T, 
                    digits = 3,
                    linesep = '',
                    caption = 'Data generating process: Fourth- to sixth-lowest RMSE methods by DGP for Monte Carlo simulations using the first 20 datasets from Dorie et al. (2019), 10 replications each.')

@@

While the AIPW, TMLE, and G-computation with a SuperLearner may be the top performing methods overall, this does not mean that there are some data-generating processes (DGPs) where some other method may do better. Across the the 20 datasets, @dorie_2019_automated vary five DGP characteristics: degree of nonlinearity, the percentage treated, overlap for the treatment group, alignment (correspondence in variables used to generate the exposure and response models), and treatment effect heterogeneity. To test how the methods perform across different values of these characteristics, I begin with the 200 simulations from the 20 datasets used in the main results (Figures \@ref(fig:dorie-results-bias) and \@ref(fig:dorie-results-rmse)). I limit datasets to those generated by a particular value of a DGP characteristic, and I then calculate RMSE for each method for only these datasets. For example, alignment is 0 for datasets 8 and 16 (meaning there is 0 correlation between the terms included in the treatment and outcome models), so for this value of the DGP RMSE is calculated only for those two datasets.

Tables \@ref(tab:dgp) and \@ref(tab:dgp-2) shows the six top-performing methods by lowest RMSE for each each value of the DGP characteristics. Across DGPs, the same methods dominate as in the full range of simulations: the SuperLearner with AIPW, TMLE, and G-computation. In fact, these three methods take the top three spots for *every* DGP variation in the simulations.

Table \@ref(tab:dgp-2) shows the fourth- to sixth-lowest RMSE methods for each DGP variation. GRF with AIPW, G-computation, and TMLE take most of these spots, but the Lin estimator and OLS appear a few times.

@@

>>>>

abstract: "Double robust methods for flexible covariate adjustment in causal inference have proliferated in recent years. Despite their apparent advantages, these methods are rarely used by social scientists. It is also unclear whether these methods actually outperform more traditional methods in finite samples. This paper has two aims: It is a guide to some of the latest methods in double robust, flexible covariate adjustment using machine learning, and it compares these methods to more traditional statistical methods and flexible \"single robust\" methods using simulated, cross-sectional data where the treatment effect is known. Double robust methods covered include Augmented Inverse Probability Weighting (AIPW), Targeted Maximum Likelihood Estimation (TMLE), and Double/Debiased Machine Learning (DML). Results suggest that some of these methods do outperform traditional methods in a wide range of simulations, but not dramatically. The top performers are TMLE and AIPW in conjunction with flexible machine learning estimators, but G-computation with the same flexible estimators obtains almost identical results, and standard regression methods have only slightly higher bias. Researchers should opt for estimators that are robust to heterogeneous treatment effects, regardless of whether they are double robust."

@@

library(SuperLearner)
library(broom)
library(knitr)
library(kableExtra)
library(here)
library(patchwork)
library(DoubleML)
library(mlr3)
library(mlr3pipelines)
library(mlr3learners)
library(mlr3extralearners)
library(haven)
library(tidyverse)

@@

lm_lin_continuous <- function (formula, covariates, data, weights, subset, clusters, 
  se_type = NULL, ci = TRUE, alpha = 0.05, return_vcov = TRUE, 
  try_cholesky = FALSE) 
{
  if (length(all.vars(f_rhs(formula))) > 1) {
    stop("The `formula` argument, `", format(formula), "`, must only have the ", 
      "treatment variable on the right-hand side of the formula. Covariates ", 
      "should be specified in the `covariates` argument like so:\n`covariates = ", 
      paste0("~ ", paste(all.vars(f_rhs(formula))[-1], 
        sep = " + ")), "`.", "\n\n See ?lm_lin.")
  if (!inherits(covariates, "formula")) {
    stop("The `covariates` argument must be specified as a formula:\n", 
      "You passed an object of class ", class(covariates))
  cov_terms <- terms(covariates)
  if (attr(cov_terms, "response") != 0) {
    stop("Must not specify a response variable in `covariates`` formula.\n", 
      "`covariates` must be a right-sided formula, such as '~ x1 + x2 + x3'")
  if (length(attr(cov_terms, "order")) == 0) {
    stop("`covariates` must have a variable on the right-hand side, not 0 or 1")
  full_formula <- update(formula, reformulate(c(".", labels(cov_terms))))
  datargs <- enquos(formula = full_formula, weights = weights, 
    subset = subset, cluster = clusters)
  data <- enquo(data)
  model_data <- clean_model_data(data = data, datargs)
  outcome <- as.matrix(model_data$outcome)
  n <- nrow(outcome)
  design_matrix <- model_data$design_matrix
  weights <- model_data$weights
  cluster <- model_data$cluster
  has_intercept <- attr(terms(formula), "intercept")
  treat_col <- which(attr(design_matrix, "assign") == 1)
  treatment <- design_matrix[, treat_col, drop = FALSE]
  design_mat_treatment <- colnames(design_matrix)[treat_col]
  # if (any(!(treatment %in% c(0, 1)))) {
  #   vals <- sort(unique(treatment))
  #   if (has_intercept) 
  #     vals <- vals[-1]
  #   n_treats <- length(vals)
  #   names(vals) <- paste0(colnames(design_matrix)[treat_col], 
  #     vals)
  #   treatment <- outer(drop(treatment), vals, function(x, 
  #     y) as.numeric(x == y))
  demeaned_covars <- design_matrix[, setdiff(colnames(design_matrix), 
    c(design_mat_treatment, "(Intercept)")), drop = FALSE]
  if (is.numeric(weights)) {
    center <- apply(demeaned_covars, 2, weighted.mean, weights)
  else {
    center <- colMeans(demeaned_covars)
  demeaned_covars <- sweep(demeaned_covars, 2, center)
  original_covar_names <- colnames(demeaned_covars)
  colnames(demeaned_covars) <- paste0(ifelse(grepl("\\:|(^.+\\()", 
    colnames(demeaned_covars)), paste0("(", colnames(demeaned_covars), 
    ")"), colnames(demeaned_covars)), "_c")
  n_treat_cols <- ncol(treatment)
  n_covars <- ncol(demeaned_covars)
  n_int_covar_cols <- n_covars * (n_treat_cols)
  interacted_covars <- matrix(0, nrow = n, ncol = n_int_covar_cols)
  interacted_covars_names <- character(n_int_covar_cols)
  for (i in 1:n_covars) {
    covar_name <- colnames(demeaned_covars)[i]
    cols <- (i - 1) * n_treat_cols + (1:n_treat_cols)
    interacted_covars[, cols] <- treatment * demeaned_covars[, 
      i]
    interacted_covars_names[cols] <- paste0(colnames(treatment), 
      ":", covar_name)
  colnames(interacted_covars) <- interacted_covars_names
  if (has_intercept) {
    X <- cbind(matrix(1, nrow = n, ncol = 1, dimnames = list(NULL, 
      "(Intercept)")), treatment, demeaned_covars, interacted_covars)
  else {
    if (n_treat_cols == 1) {
      X <- cbind(treatment, demeaned_covars, interacted_covars)
    else {
      X <- cbind(treatment, interacted_covars)
  return_list <- lm_robust_fit(y = outcome, X = X, weights = weights, 
    cluster = cluster, ci = ci, se_type = se_type, alpha = alpha, 
    return_vcov = return_vcov, try_cholesky = try_cholesky, 
    has_int = has_intercept, iv_stage = list(0))
  return_list <- lm_return(return_list, model_data = model_data, 
    formula = formula)
  return_list[["scaled_center"]] <- center
  setNames(return_list[["scaled_center"]], original_covar_names)
  return_list[["call"]] <- match.call()
  return(return_list)

environment(lm_lin_continuous) <- asNamespace('estimatr')
assignInNamespace("lm_lin", lm_lin_continuous, ns = "estimatr")

@@

In causal inference, functional form misspecification of underlying models can bias estimates of treatment effects [@hernan_2020_causal; @morgan_2015_counterfactuals]. There have been two important advances that attempt to overcome this. First, methodologists have developed machine learning methods that allow greater flexibility in estimation, adjusting for covariates in data-driven, complex ways [@balzer_2021_invited; @brand_2023_recent]. The second development is double robust methods [@bang_2005_doubly; @kang_2007_demystifying], which estimate two models: one for treatment exposure and another for the outcome. These models are robust to misspecification of either one of these "nuisance" models.

@@

Second, this paper evaluates these methods, testing them on simulations from @dorie_2019_automated that cover a range of data-generating processes where ignorability holds -- meaning there are no unmeasured confounders. In the simulations, double robust methods are compared to "single robust" methods (i.e., ones with one nuisance model), including traditional or simpler statistical methods commonly used by social scientists: ordinary least squares (OLS) regression, matching on propensity scores estimated from logistic regression (PSM), and inverse probability weighting (IPW). They are also compared to two more flexible methods that may overcome the misspecification issues that motivate double robust methods: G-computation and the Lin estimator. G-computation models the outcome in separate models for treated and untreated units, then predicts the average difference between these for the full sample [@robins_1986_new]. The Lin estimator interacts treatment with mean-centered covariates in an OLS regression [@lin_2013_agnostic].

Results show that some double robust methods outperform traditional statistical methods, but not dramatically. AIPW and TMLE perform the best, at least when used in conjunction with flexible machine learning algorithms, while DML does slightly worse than traditional methods, at least in its partially linear form that is not robust to heterogeneous treatment effects. G-computation with flexible machine learning performs as well as the lowest-error double robust methods. With its still relatively low error and much faster computation time, OLS remains a sensible choice as an estimator, and the Lin estimator -- which is as quick as standard OLS regression -- performs only slightly worse than the top machine learning methods. Due to the strong performance of G-computation and the Lin estimator, results suggest that using estimators that are robust to heterogeneous treatment effects may be more important than double robustness.

The next section reviews literature on and motivation for double robust methods. This is followed by an introduction of notation and assumptions, while the following section gives an overview of the double robust methods covered in this paper. I next present results of the simulations, then conclude. In the Appendix, I provide simple R code to implement the three double robust methods.

@@

# Background

@@

<!-- Although previous work has compared different methods for covariate adjustment, there has not yet been a thorough comparison and evaluation of the recent and popular double robust methods of AIPW, TMLE, and DML for a social scientist audience.  -->

@@

# Overview of Techniques  
Each of the methods reviewed in this paper can be used with a variety of estimation techniques, including both traditional statistical methods and flexible machine learning algorithms. Each involves a model for the outcome and another for the treatment exposure, but choice of estimator for these two models is left to the discretion of the user. Double robust methods are distinct in the ways these estimated models relate and are combined into a final estimate of the desired estimand.

This section provides some intuition for the mathematical theory behind each method. In the Appendix, I provide R code to simply implement these methods. This is the code used to evaluate these methods later in this paper, and it represents basic implementations of these methods. However, for researchers wishing to put these methods into practice, using a dedicated R package for each method is probably a better idea. These packages have many flexible options, such as accounting for complex survey design, targeting estimands besides the ATE, and integrating with a variety of estimation techniques.^[In the Supplementary Material full tables of results, I present the results of using two packages to estimate the full set of simulations shown below. These include AIPW in conjunction with generalized random forests (GRF) using the `grf` package [@tibshirani_2024_grf] and a partially linear model for DML with a SuperLearner that, like below, harnesses ensemble learning with GLM, glmnet, and XGBoost, using the `DoubleML` package [@bach_2022_doubleml]. Results are fairly similar to my own R code, though computation time is markedly faster.]

@@

Extending and improving previous double robust methods, @vanderlaan_2006_targeted first proposed TMLE using a parametric framework and the efficient influence curve [@hines_2022_demystifying] to obtain estimates and standard errors. Van der Laan has gone on to collaborate on both a gentle introduction [@gruber_2009_targeted], two textbooks [@vanderlaan_2011_targeted; @vanderlaan_2018_targeted], and an R package [@gruber_2012_tmle] for implementing the method. @schuler_2017_targeted and @luque-fernandez_2018_targeted provide introductions for epidemiologists.

@@

In the next step, we estimate fluctuation parameters $\epsilon = (\epsilon_0, \epsilon_1)$ through maximum likelihood of the following logistic regression with fixed intercept $\text{logit}(\mu_{di})$: 
<!-- $$E(Y=1 \mid D, \mathbf X) = \text{logit}(\mu_{di}) + \epsilon H_{di}$$ -->
$$\text{logit}[E(Y=1 \mid D, \mathbf X)] = \text{logit}(\hat \mu_{di}) + \epsilon_0 H_{0i} + \epsilon_1 H_{1i}.$$

@@

For each individual in the sample, this estimator calculates two quantities: The treated potential outcome

@@

and the control potential outcome

\begin{equation}
\hat Y_{0i} = \frac{(1-D_i)(Y_i - \hat \mu_0 (X_i))}{1-\hat \pi(X_i)} + \hat \mu_0(X_i).
(\#eq:control-aipw)
\end{equation}

Let's focus on the treated model, Equation \@ref(eq:treated-aipw). First, assume that the outcome model $\mu_1(X_i)$ is *correctly* specified and the exposure model $\pi(X_i)$ is *incorrectly* specified. Let's also assume (for now) that we're dealing with a treated unit, i.e. $D_i = 1$. Then
$$E[\hat \mu_1 (X_i)] = E[Y_1 \mid X_i].$$
This means that the model for the outcome and the observed outcome for this treated unit are equal in expectation, so
$$E[Y_i - \hat \mu_1(X_i)] = 0.$$
Plugging into Equation \@ref(eq:treated-aipw), we get 
$$E[\hat Y_{1i}] = 0 + \hat \mu_1(X_i) = \hat \mu_1(X_i).$$
So the model relies *only* on the outcome model. The incorrectly specified exposure model completely disappears from the equation. If we're dealing with a control unit, we get the same result, plugging $D_i=0$ into the same Equation \@ref(eq:treated-aipw):
$$\hat Y_{1i} = \frac{0(Y_i - \hat \mu_1 (X_i))}{\hat \pi (X_i)} + \hat \mu_1(X_i) = \hat \mu_1(X_i).$$

Now, what if the *exposure* model $\pi(X_i)$ is correctly specified and the outcome model $\mu_1(X)$ is incorrect? First, we use algebra to rewrite Equation \@ref(eq:treated-aipw) for the treated outcome:
\begin{align}
\hat Y_{1i}& = \frac{D_i(Y_i - \hat \mu_1 (X_i))}{\hat \pi (X_i)} + \hat \mu_1(X_i) \nonumber \\
&= \frac{D_iY_i}{\hat \pi (X_i)} - \frac{D_i\hat \mu_1 (X_i)}{\hat \pi (X_i)} + \frac{\hat \pi (X_i)\hat \mu_1(X_i)}{\hat \pi (X_i)} \nonumber \\
& = \frac{D_iY_i}{\hat \pi (X_i)} - \left( \frac{D_i - \hat \pi(X_i)}{\hat \pi (X_i)}\right) \hat \mu_1(X_i). 
(\#eq:exposure-correct)
\end{align}
Since the exposure model is correctly specified, we have $D_i = \hat \pi(X_i)$ on average, so
$$E[D_i - \hat \pi(X_i)] = 0.$$
This means that the second term in Equation \@ref(eq:exposure-correct) is 0 in expectation, so
$$E[\hat Y_{1i}]= E \left [ \frac{D_iY_i}{\hat \pi (X_i)}\right].$$
This shows that when the exposure model is correct, then the estimator depends *only* on the exposure model. We can make similar arguments for the control model for $\hat Y_{0i}$ in Equation \@ref(eq:control-aipw).

This demonstration shows that this estimator achieves double robustness: The estimator is robust to misspecification of either the exposure or the outcome model (but not both). The other two double robust methods considered in this paper can be shown to have the same property, but demonstrating this is more complicated. Interested readers can refer to the references cited above.

@@

```{r dml-package, eval = F}

@@

In 2016, the Atlantic Causal Inference Conference hosted a competition for causal inference methods that adjust on observables. @dorie_2019_automated published the results of this competition, along with the data used. Below, I test double robust methods on the 20 data generating processes (DGPs) reserved for the "do-it-yourself" part of the competition. The data represent a hypothetical twins study investigating the impact of birth weight on IQ. The data have 4,802 observations and 52 covariates. Treatment is binary and the outcome is continuous. In all DGPs, ignorability holds (all potential confounders are observed), but the authors vary the following: (1) degree of nonlinearity, (2) percentage of treated, (3) overlap for the treatment group, (4) alignment (correspondence in variables used to generate the assignment mechanism and the response surface), and (5) treatment effect heterogeneity.

@@

The second set of single robust methods are two-model methods, which estimate separate models for treated and untreated units. In theory these could solve the misspecification problem that double robust methods are meant to solve, but they could still suffer from regularization bias. G-computation -- also called regression imputation -- uses some estimation technique to predict outcomes under treatment and control for each unit in the dataset [@robins_1986_new; @snowden_2011_implementation]. This underlying technique can be anything from a traditional method such as OLS regression to a flexible machine learning method. The ATE estimate is the difference in the average prediction under treatment and the average prediction under control:

$$\begin{aligned}
\hat \tau_{gcomp} = &\frac{1}{n} \sum_{i=1}^n \left[\hat \mu_1 (\mathbf X_i) - \hat \mu_0 (\mathbf X_i) \right].
\end{aligned}$$

The second two-model method is the Lin estimator [@lin_2013_agnostic]. This method aims to solve issues with the bias induced by OLS regression in a randomization framework by interacting the treatment indicator with mean-centered covariates. @hazlett_2024_demystifying show that this method is equivalent to estimating two separate OLS regression models for the treated and control units -- i.e., G-computation with each of the underlying models estimated by OLS regression.

@@

I use 10 simulations of each of the 20 DGPs, resulting in 200 data sets. I then calculate bias, percent bias (the estimator's bias as a percentage of its standard error), root mean squared error (rmse), and median absolute error (mae). I also present the number of datasets for which the method fails and the median computation time for each data set, in seconds.^[Simulations were run on a 2020 MacBook Pro laptop computer with a 2 GHz Quad-Core Intel Core i5 Processor and 16 GB of memory.] In the main text I present average bias and RMSE, while the Supplementary Material contains tables with full results.

@@

```{r dorie-results-bias,  fig.cap = 'Bias of Monte Carlo simulations using the first 20 DGPs from Dorie et al. (2019), 10 replications each. Points represent the bias of estimates on individual datasets, while box plots show the median, 25th and 75th percentiles, and whiskers extending to 1.5 times the interquartile range. Colors correspond to method (in which multiple estimators may be used).'}
addline_format <- function(x,...){
    gsub('\\s','\n',x)
}

@@

```{r dorie-results-rmse, fig.cap = 'Root mean squared error and bias for Monte Carlo simulations using the first 20 DGPs from Dorie et al. (2019), 10 replications each. Values greater in absolute value than 4 are plotted at 4 and labeled with their actual value.'}

@@

```{r dorie-linear-bias,  fig.cap = 'Bias of Monte Carlo simulations using the two datasets from Dorie et al. (2019), with linear data generating processes, 100 replications each. Points represent the bias of estimates on individual datasets, while box plots show the median, 25th and 75th percentiles, and whiskers extending to 1.5 times the interquartile range. Colors correspond to method (in which multiple estimators may be used).'}
sim_results %>% 
  filter(set == 'linear') %>%
  mutate(method_estimator = str_replace(method_estimator, 'superlearner', 'superl.')) %>%
  mutate(method_estimator = factor(method_estimator, levels = unique({.$method_estimator}))) %>%
  #filter(method %in% c('ols', 'psm', 'aipw')) %>%
  ggplot(aes(x = method_estimator, y = bias)) +
  geom_jitter(alpha = .2, width = .3, aes(color = method)) +
  geom_boxplot(alpha = .5, outlier.alpha = 0) +
  geom_hline(yintercept = 0) +
  #facet_wrap(~method, scales = 'free_x') +
  labs(x = '', title = 'Linear DGPs') +
  scale_y_continuous(trans=ggallin::pseudolog10_trans, breaks = c(-20, -10, -5, -1, 0, 1, 5)) +
  # scale_x_discrete(labels=function(x){sub(", ", "\n", x)}) +
  theme(axis.text.x=element_text(angle=90, hjust=0.95,vjust=0.2),
        legend.position = 'none')

@@

bind_rows(dgp_list) %>%
  select(-c(`Fourth-lowest`:`Sixth-lowest`)) %>%
  kableExtra::kable(booktabs = T, 
                    digits = 3,
                    linesep = '',
                    caption = 'Data generating process: Three lowest RMSE methods by DGP for Monte Carlo simulations using the first 20 DGPs from Dorie et al. (2019), 10 replications each.')
```

```{r dgp-2}
bind_rows(dgp_list) %>%
  select(-c(`Lowest RMSE`:`Third-lowest`)) %>%
  kableExtra::kable(booktabs = T, 
                    digits = 3,
                    linesep = '',
                    caption = 'Data generating process: Fourth- to sixth-lowest RMSE methods by DGP for Monte Carlo simulations using the first 20 DGPs from Dorie et al. (2019), 10 replications each.')

@@

While the AIPW, TMLE, and G-computation with a SuperLearner may be the top performing methods overall, this does not mean that there are some DGPs where some other method may do better. Across the 20 DGPs, @dorie_2019_automated vary five characteristics: degree of nonlinearity, the percentage treated, overlap for the treatment group, alignment (correspondence in variables used to generate the exposure and response models), and treatment effect heterogeneity. To test how the methods perform across different values of these characteristics, I begin with the 200 simulations from the 20 DGPs used in the main results (Figures \@ref(fig:dorie-results-bias) and \@ref(fig:dorie-results-rmse)). I limit datasets to those generated by a particular value of a DGP characteristic, and I then calculate RMSE for each method for only these datasets. For example, alignment is 0 for datasets 8 and 16 (meaning there is 0 correlation between the terms included in the treatment and outcome models), so for this value RMSE is calculated only for datasets generated from those two DGPs.

Tables \@ref(tab:dgp) and \@ref(tab:dgp-2) shows the six top-performing methods by lowest RMSE for each each value of the DGP characteristics. (The Supplementary Material presents a table with full results for all DGPs in the simulations.) Across DGPs, the same methods dominate as in the full range of simulations: the SuperLearner with AIPW, TMLE, and G-computation. In fact, these three methods take the top three spots for *every* DGP variation in the simulations.

Table \@ref(tab:dgp-2) shows the fourth- to sixth-lowest RMSE methods for each DGP variation. GRF with AIPW, G-computation, and TMLE take most of these spots, but the Lin estimator and OLS appear a few times. Notably, in the presence of high treatment effect heterogeneity, flexible machine learning methods take the top six spots, and not all of these are double robust. For the DGPs, allowing for treatment effects to vary appears more important than double robustness. This is supported by the performance of the Lin estimator, which takes the seventh spot (as shown in the Supplementary Material). This estimator is relatively inflexible but robust to heterogeneous effects. DML, which as implemented here is not robust to heterogeneous treatment effects, performs worse than OLS regression.

@@

```{r hetro-rmse, eval = F, fig.height = 7, fig.cap = 'Treatment Effect Heterogeneity: Root mean squared error and bias for Monte Carlo simulations from the first 20 DGPs in Dorie et al. (2019), 10 replications each. Values greater in absolute value than 4 are plotted at 4 and labeled with their actual value.'}
